---
title: "Reproducible_research_proj_1"
author: "Yiran Chen"
date: "2025-04-11"
output: html_document
---

## Load the data
```{r loaddata}
data <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
```{r meanstep}
# Remove NA data
clean_data <- na.omit(data)

# Calculate the total number of steps taken per day
total_steps_daily <- aggregate(steps ~ date, data = clean_data, sum)

# Calculate the mean
mean_total_step <- mean(total_steps_daily$steps)

```

So the mean of total number of steps taken per day is `r mean_total_step`.

See the histogram below.
```{r hist}
hist(total_steps_daily$steps,
     main = "Histogram of Total Steps per Day",
     xlab = "Total Steps per Day",
     col = "skyblue",
     border = "white")
```

The mean and median total steps per day calculation is shown below.
```{r mean-median-steps}
median_total_step <- median(total_steps_daily$steps)

cat("Mean of total steps per day:", mean_total_step, "\n")
cat("Median of total steps per day:", median_total_step, "\n")
```

## What is the average daily activity pattern?
```{r activity-pattern}
avg_steps_per_interval <- aggregate(steps ~ interval, data = data, FUN = mean, na.rm = TRUE)

# Plot the time series
plot(avg_steps_per_interval$interval,
     avg_steps_per_interval$steps,
     type = "l",
     xlab = "5-minute Interval",
     ylab = "Average Number of Steps",
     main = "Average Daily Activity Pattern")

# Find the max value of steps taken for the 5-minute interval
max_interval <- avg_steps_per_interval[which.max(avg_steps_per_interval$steps), ]

# Print result
cat("The 5-minute interval with the maximum average number of steps is:", 
    max_interval$interval, 
    "with an average of", 
    max_interval$steps, 
    "steps.\n")

```

## Imputing missing values
```{r missingvals}
# Total number of NA values in 'steps'
total_na <- sum(is.na(data$steps))
cat("Total number of missing values in 'steps':", total_na, "\n")

# Fill in the missing values with the mean of that 5-min interval
interval_means <- aggregate(steps ~ interval, data = data, FUN = mean, na.rm = TRUE)
data_imputed <- merge(data, interval_means, by = "interval", suffixes = c("", ".mean"))

data_imputed$steps[is.na(data_imputed$steps)] <- data_imputed$steps.mean
data_imputed$steps.mean <- NULL

# Recalculate total steps per day with imputed data
steps_daily_imputed <- aggregate(steps ~ date, data = data_imputed, sum)

# Make plot
hist(steps_daily_imputed$steps,
     main = "Histogram of Total Steps per Day (Imputed Data)",
     xlab = "Total Steps per Day",
     col = "lightgreen",
     border = "white")
```

After imputation, recalculate the mean and median as follows.
```{r mean-median-recal}
mean_imputed <- mean(steps_daily_imputed$steps)
median_imputed <- median(steps_daily_imputed$steps)

cat("Mean (imputed):", mean_imputed, "\n")
cat("Median (imputed):", median_imputed, "\n")

```

From earlier calculation, we know that the mean was `r mean_total_step` and median `r median_total_step`. So we can see the difference caused by missing values.

## Differences in activity patterns between weekdays and weekends?
```{r add-weekday-weekend}
data_imputed$date <- as.Date(data_imputed$date)
data_imputed$day_type <- ifelse(weekdays(data_imputed$date)%in% c("Saturday","Sunday"),"weekend","weekday")
data_imputed$day_type <- factor(data_imputed$day_type, levels = c("weekday","weekend"))
```

```{r recal-mean}
library(dplyr)
interval_daytype_avg <- data_imputed %>%
  group_by(interval, day_type) %>%
  summarise(average_steps = mean(steps), .groups = "drop")
```

See the plot below.
```{r weekdays-vs-weekends-plot}
library(ggplot2)
ggplot(interval_daytype_avg, aes(x = interval, y = average_steps)) +
  geom_line() +
  facet_wrap(~ day_type, ncol = 1) +
  labs(title = "Activity Patterns: Weekday vs. Weekend",
       x = "5-minute Interval",
       y = "Average Steps") +
  theme_minimal()
```